---
title: "Main"
author: "Daniel Vasilevko"
date: "2026-05-02"
output: pdf_document
---

```{r}
library(dplyr)
library(tidyr)
library(tidycensus)
library(ggplot2)
library(viridis)

library(tigris)
library(sf)
```

```{r}
departments <- read.csv("departments2.csv")
diagnosis <- read.csv("diagnosis2.csv")
encounters <- read.csv("encounters2.csv")
patients2 <- read.csv("patients2.csv")
providers <- read.csv("providers2.csv" )
social_det <- read.csv("social_determinant2.csv")
tigercensus <- read.csv("tigercensuscodes2.csv")

combined <- read.csv("df.csv")
```

# Making Geo DATA
```{r}
geo_data <- read.csv("geographic_data.csv")

Unemployemnt <- read.csv("Unemployment_Rate.csv")

clean_join <- geo_data %>%
  inner_join(
    Unemployemnt,
    by = c("GEOID" = "GEOID")
  )

clean_join <- clean_join %>%
  mutate(
    Unemployment.Rate = ifelse(Unemployment.Rate == "-100.00%", NA, Unemployment.Rate)
  )

#write.csv(clean_join, "geographic_data.csv", row.names = FALSE)

geo_data <- read.csv("geographic_data.csv")

encounters

```

```{r}
combined <- read.csv("df.csv")

n_distinct(patients2$DurableKey)

combined
```

```{r}
df_sorted <- combined %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%y")) %>%
  arrange(PatientDurableKey, Date)

df_sorted
```

# Generalizing The Diagnosis Values
```{r}
diagnosis2 <- diagnosis %>%
  mutate(DiagnosisValue = substr(DiagnosisValue, 1, 3))

diagnosis2 <- diagnosis2 %>%
  mutate(
    ICD_Category = case_when(
      substr(DiagnosisValue, 1, 1) %in% c("A", "B") ~ "Infectious and parasitic diseases",
      substr(DiagnosisValue, 1, 1) == "C" ~ "Neoplasms (cancers)",
      substr(DiagnosisValue, 1, 1) == "D" ~ "Neoplasms, Blood, Blood-forming Organs",
      substr(DiagnosisValue, 1, 1) == "E" ~ "Endocrine, nutritional, metabolic",
      substr(DiagnosisValue, 1, 1) == "F" ~ "Mental and behavioral disorders",
      substr(DiagnosisValue, 1, 1) == "G" ~ "Nervous system diseases",
      substr(DiagnosisValue, 1, 1) == "H" ~ "Eye and ear diseases",
      substr(DiagnosisValue, 1, 1) == "I" ~ "Circulatory system",
      substr(DiagnosisValue, 1, 1) == "J" ~ "Respiratory system",
      substr(DiagnosisValue, 1, 1) == "K" ~ "Digestive system",
      substr(DiagnosisValue, 1, 1) == "L" ~ "Skin diseases",
      substr(DiagnosisValue, 1, 1) == "M" ~ "Musculoskeletal system",
      substr(DiagnosisValue, 1, 1) == "N" ~ "Genitourinary system",
      substr(DiagnosisValue, 1, 1) == "O" ~ "Pregnancy and childbirth",
      substr(DiagnosisValue, 1, 1) == "P" ~ "Perinatal conditions",
      substr(DiagnosisValue, 1, 1) == "Q" ~ "Congenital conditions",
      substr(DiagnosisValue, 1, 1) == "R" ~ "Symptoms / abnormal findings",
      substr(DiagnosisValue, 1, 1) %in% c("S", "T") ~ "Injuries and external causes",
      substr(DiagnosisValue, 1, 1) %in% c("V", "W", "X", "Y") ~ "External causes (accidents, injuries)",
      substr(DiagnosisValue, 1, 1) == "Z" ~ "Factors influencing health",
      TRUE ~ "Other / Unknown"
    )
  )

diagnosis2
```

# Adding ICD To Each Diagnosis Key Journey Data 
```{r}
diagnosis_times <- read.csv("diagnosis_times.csv")

diagnosis_times_large <- read.csv("diagnosis_times_large.csv")

diagnosis3 <- diagnosis2 %>%
  select(DiagnosisKey, ICD_Category)

diagnosis3

df_joined <- diagnosis_times_large %>%
  left_join(
    diagnosis3,
    by = c("PrimaryDiagnosisKey" = "DiagnosisKey")
  )

df_joined

dx_with_icd <- df_joined %>%
  group_by(PatientDurableKey, PrimaryDiagnosisKey) %>%
  summarise(
    num_encounters = first(num_encounters),
    first_date = first(first_date),
    last_date = first(last_date),
    ICD_Categories = list(unique(ICD_Category)),
    .groups = "drop"
  )



```

# Adding Location to Every Patient Key in Jounrey Data
```{r}
patients2_subset <- patients2 %>%
  select(DurableKey, CensusBlockGroupFipsCode)

patients2_subset

dx_joined <- dx_with_icd_clean %>%
  left_join(
    patients2_subset,
    by = c("PatientDurableKey" = "DurableKey")
  )

dx_joined %>%
  summarise(
    total = n(),
    unspecified = sum(CensusBlockGroupFipsCode == "*Unspecified", na.rm = TRUE),
    percent_unspecified = 100 * unspecified / total
  )

dx_joined
```

# Adding Information About Location in Journey Data
```{r}
geo_data <- read.csv("geographic_data.csv")

geo_data <- geo_data %>%
  mutate(GEOID = as.character(GEOID))

dx_geo_joined <- dx_joined %>%
  left_join(
    geo_data,
    by = c("CensusBlockGroupFipsCode" = "GEOID")
  )

dx_geo_joined
```

# Large Joureny
```{r}
df_joined2 <- diagnosis_times %>%
  left_join(
    diagnosis3,
    by = c("PrimaryDiagnosisKey" = "DiagnosisKey")
  )

dx_with_icd2 <- df_joined2 %>%
  group_by(PatientDurableKey, PrimaryDiagnosisKey) %>%
  summarise(
    num_encounters = first(num_encounters),
    first_date = first(first_date),
    last_date = first(last_date),
    ICD_Categories = list(unique(ICD_Category)),
    .groups = "drop"
  )

dx_with_icd2 <- dx_with_icd2 %>%
  mutate(ICD_Categories = sapply(ICD_Categories, paste, collapse = ", "))

dx_with_icd2 <- dx_with_icd2 %>%
  left_join(
    patients2_subset,
    by = c("PatientDurableKey" = "DurableKey")
  )

large_journey <- dx_with_icd2 %>%
  left_join(
    geo_data,
    by = c("CensusBlockGroupFipsCode" = "GEOID")
  )

```

# Graphing Car Access Affect on Num Encounters
```{r}
journey_data <- read.csv("journey_data.csv")

patient_summary <- journey_data %>%
  group_by(PatientDurableKey) %>%
  summarise(
    total_encounters = sum(num_encounters, na.rm = TRUE),
    No.Car.RATE = first(No.Car.RATE),  # assumes one geo per patient
    .groups = "drop"
  )

patient_summary_clean <- patient_summary %>%
  filter(
    !is.na(No.Car.RATE),
    !is.na(total_encounters)
  )

patient_summary_clean <- patient_summary_clean %>%
  mutate(
    No.Car.RATE = as.numeric(gsub("%", "", No.Car.RATE)) / 100
  )

model <- lm(total_encounters ~ No.Car.RATE, data = patient_summary_clean)
summary(model)

ggplot(patient_summary_clean, aes(x = No.Car.RATE, y = total_encounters)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "No Car Rate",
    y = "Total Encounters",
    title = "Transportation Barriers vs Healthcare Utilization"
  ) +
  theme_classic()
```

# Graphing Median Income Affect on Num Encounters
```{r}
journey_data <- journey_data %>%
  mutate(
    Median.Household.Income = as.numeric(Median.Household.Income)
  )

patient_income <- journey_data %>%
  group_by(PatientDurableKey) %>%
  summarise(
    total_encounters = sum(num_encounters, na.rm = TRUE),
    income = first(Median.Household.Income),
    .groups = "drop"
  ) %>%
  filter(!is.na(income))


model1 <- lm(total_encounters ~ income, data = patient_income)
summary(model1)

library(ggplot2)

ggplot(patient_income, aes(x = income, y = total_encounters)) +
  geom_point(alpha = 0.4, color = "darkturquoise") +
  labs(
    x = "Median Household Income (Per Census Block)",
    y = "Total Encounters"  )+
  theme_classic()+
  theme(
  plot.background = element_rect(fill = "#fafafa", color = NA),
  panel.background = element_rect(fill = "#fafafa", color = NA),

  axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),

  axis.title.x = element_text(size = 15),
  axis.title.y = element_text(size = 15))
```
```{r}
patient_income <- patient_income %>%
  mutate(
    income_group = cut(
      income,
      breaks = c(0, 20000, 40000, 60000, 80000, 100000, 120000, 150000,Inf),
      labels = c("0-20k", "20-40k", "40-65k", "60-80k", "80-100k", "100-120k","120-150k","150K+"),
      right = FALSE
    )
  )

income_summary <- patient_income %>%
  group_by(income_group) %>%
  summarise(
    avg_encounters = mean(total_encounters, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(income_summary, aes(x = income_group, y = avg_encounters)) +
  geom_col(fill = "steelblue", alpha = 0.85) +
  geom_text(
    aes(label = round(avg_encounters, 1)),
    vjust = -0.3,
    size = 4
  ) +
  labs(
    x = "Income Group",
    y = "Avg Encounters",
    title = "Healthcare Utilization by Income"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```


# Graphing Unuemployemnt Effect on Num Encounters 
```{r}
journey_data 

dx_geo_joined <- dx_geo_joined %>%
  mutate(
    Unemployment.Rate = as.numeric(gsub("%", "", Unemployment.Rate))
  )

patient_unemp <- dx_geo_joined %>%
  group_by(PatientDurableKey) %>%
  summarise(
    total_encounters = sum(num_encounters, na.rm = TRUE),
    unemployment = first(Unemployment.Rate),
    .groups = "drop"
  ) %>%
  filter(!is.na(unemployment))

model2 <- lm(total_encounters ~ unemployment, data = patient_unemp)

summary(model2)
```

# Kansas Map
```{r}
ks_counties <- counties(state = "KS", cb = TRUE, class = "sf")
ks_tracts <- tracts(state = "KS", cb = TRUE, class = "sf")

ggplot(ks_counties) +
  geom_sf(fill = "lavender", color = "black") +
  theme_void() +
  ggtitle("Kansas Counties")
```
```{r}
building_locations <- departments %>%
  distinct(CensusTract)

building_locations <- building_locations[
  substr(building_locations$CensusTract, 1, 2) == "20",
]

ks_centroids <- ks_tracts %>%
  st_centroid() %>%
  mutate(
    GEOID = as.character(GEOID),
    lon = st_coordinates(.)[,1],
    lat = st_coordinates(.)[,2]
  )

building_locations <- as.character(building_locations)

building_geo <- ks_centroids %>%
  filter(GEOID %in% building_locations)

building_geo

clean_coords <- building_geo %>%
  mutate(
    lon = st_coordinates(.)[,1],
    lat = st_coordinates(.)[,2]
  ) %>%
  st_drop_geometry() %>%
    select(GEOID, lat, lon)

clean_coords
```
```{r}
ks_map <- tracts(state = "KS", cb = TRUE, class = "sf")

coords_sf <- st_as_sf(
  clean_coords,
  coords = c("lon", "lat"),
  crs = 4326
)

ggplot() +
  geom_sf(data = ks_counties, fill = "white", color = "gray70") +
  geom_point(
    data = coords_sf,
    aes(x = st_coordinates(coords_sf)[,1],
        y = st_coordinates(coords_sf)[,2]),
    color = "deepskyblue",
    shape = 2,   
    size = 2.5,
    alpha = 0.8
  ) +
  coord_sf(
    xlim = c(-97.65, -94),
    ylim = c(37.47, 40.5),
    expand = FALSE
  ) +
  theme_void() +
  ggtitle("Kansas Census Tracts with Locations")
```

# Closest Hospital
```{r}
large_journey2 <- read.csv("large_journey_data.csv")

patients_clean <- large_journey2 %>%
  filter(!is.na(CENTLAT), !is.na(CENTLON))

patients_sf <- st_as_sf(
  patients_clean,
  coords = c("CENTLON", "CENTLAT"),
  crs = 4326
)

patients_sf <- st_transform(patients_sf, 3857)
buildings_sf <- st_transform(coords_sf, 3857)

nearest_idx <- st_nearest_feature(patients_sf, buildings_sf)
patients_sf$nearest_building <- buildings_sf$GEOID[nearest_idx]

patients_sf$distance_m <- as.numeric(
  st_distance(patients_sf, buildings_sf[nearest_idx, ], by_element = TRUE)
)

patients_sf$distance_km <- patients_sf$distance_m / 1000

Distance_journey <- st_drop_geometry(patients_sf)

patients_sf <- patients_sf %>%
  select(nearest_building, distance_m, distance_km, geometry)


```
```{r}
patients_sf <- patients_sf %>%
  mutate(
    closeness = exp(-distance_km / 50)
  )

patients_plot <- patients_sf %>%
  mutate(
    lon = st_coordinates(.)[,1],
    lat = st_coordinates(.)[,2]
  )

patients_plot <- st_transform(patients_plot, 4326)
buildings_sf  <- st_transform(buildings_sf, 4326)


# -----------------------------
# 2. Extract coordinates once
# -----------------------------
coords <- st_coordinates(patients_plot)
patients_plot$lon <- coords[,1]
patients_plot$lat <- coords[,2]


# -----------------------------
# 3. Remove duplicates / aggregate
# -----------------------------
patients_plot <- patients_plot |>
  group_by(lon, lat) |>
  summarise(closeness = mean(closeness), .groups = "drop")

patients_plot

# -----------------------------
# 4. Log-scale closeness (optional but recommended)
# -----------------------------


ggplot() +
  geom_sf(data = ks_counties, fill = "#fafafa", color = "gray70") +

  geom_bin2d(
    data = patients_plot,
    aes(x = lon, y = lat, weight = closeness),
    bins = 250
  ) +

  scale_fill_viridis_c(
    option = "inferno",
    trans = "sqrt",   # 👈 key line
    name = "Closeness"
  ) +

  geom_point(
    data = coords_sf,
    aes(x = st_coordinates(coords_sf)[,1],
        y = st_coordinates(coords_sf)[,2]),
    color = "darkturquoise",
    shape = 2,
    size = 1.5,
    alpha = 0.8
  ) +

  coord_sf(
    xlim = c(-102.5, -94),
    ylim = c(36.5, 40.5),
    expand = FALSE
  ) +

  theme_void()+
  theme(
    plot.background = element_rect(fill = "#fafafa", color = NA),
    panel.background = element_rect(fill = "#fafafa", color = NA)
  )


```
```{r}
ggplot() +
  geom_sf(data = ks_counties, fill = "white", color = "gray70") +

  geom_bin2d(
    data = patients_plot,
    aes(x = lon, y = lat, weight = closeness),
    bins = 250
  ) +

  scale_fill_viridis_c(
    option = "inferno",
    trans = "sqrt",   # 👈 key line
    name = "Closeness"
  ) +

  geom_point(
    data = coords_sf,
    aes(x = st_coordinates(coords_sf)[,1],
        y = st_coordinates(coords_sf)[,2]),
    color = "darkturquoise",
    shape = 2,
    size = 3.5,
    alpha = 0.8
  ) +

  coord_sf(
    xlim = c(-98, -94),
    ylim = c(38.2, 40.5),
    expand = FALSE
  ) +

  theme_void()
```

# Encounters Based on Distance
```{r}
patient_distance <- Distance_journey %>%
  group_by(PatientDurableKey) %>%
  summarise(
    total_encounters = sum(num_encounters, na.rm = TRUE),
    distance_km = first(distance_km),
    .groups = "drop"
  ) %>%
  filter(!is.na(distance_km))

model_dist <- lm(total_encounters ~ distance_km, data = patient_distance)
summary(model_dist)


ggplot(patient_distance, aes(x = distance_km, y = total_encounters)) +
  geom_point(alpha = 0.4, color = "darkturquoise") +
  labs(
    x = "Patient Distance to Nearest Facility (km)",
    y = "Total Encounters per Patient"
  ) +
  theme_classic()+
  theme(
  plot.background = element_rect(fill = "#fafafa", color = NA),
  panel.background = element_rect(fill = "#fafafa", color = NA),

  axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14),

  axis.title.x = element_text(size = 16),
  axis.title.y = element_text(size = 16)
)


```

# Distribution of Multiple Diagnosis
```{r}
multiple_diagnosis <- read.csv("large_journey_data.csv")

multiple_diagnosis

ggplot(multiple_diagnosis, aes(x = num_encounters)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(
    x = "Number of Encounters (per Patient–Diagnosis)",
    y = "Count of Patient–Diagnosis Pairs",
    title = "Distribution of Encounters per Diagnosis Key"
  ) +
  theme_classic()

```

# Transporation Survey
```{r}
unique(social_det$DisplayName)

transport <- social_det[social_det$DisplayName=="In the past 12 months, has lack of transportation kept you from medical appointments or from getting medications?",]

patient_answers1 <- transport %>%
  group_by(PatientDurableKey, AnswerText) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(PatientDurableKey, desc(n)) %>%
  group_by(PatientDurableKey) %>%
  slice(1) %>%
  ungroup() %>%
  select(PatientDurableKey, AnswerText)

patient_totals <- large_journey2 %>%
  group_by(PatientDurableKey) %>%
  summarise(
    total_encounters = sum(num_encounters, na.rm = TRUE),
    .groups = "drop"
  )

patient_summary <- patient_answers %>%
  left_join(patient_totals, by = "PatientDurableKey")

patient_summary

model_transport_survey <- lm(total_encounters ~ AnswerText, data = patient_summary)
summary(model_transport_survey)

patient_summary %>%
  group_by(AnswerText) %>%
  summarise(
    avg_encounters = mean(total_encounters, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = AnswerText, y = avg_encounters, fill = AnswerText)) +
  geom_col(alpha = 0.9) +

  # consistent color palette
  scale_fill_manual(values = c(
    "No" = "darkturquoise",
    "Yes" = "#00BFC4"   # slightly deeper teal for contrast
  )) +

  geom_text(
    aes(label = round(avg_encounters, 1)),
    vjust = -0.3,
    size = 5
  ) +

  labs(
    x = "Transportation Need",
    y = "Average Encounters per Patient"
  ) +

  theme_classic() +
  theme(
    plot.background = element_rect(fill = "#fafafa", color = NA),
    panel.background = element_rect(fill = "#fafafa", color = NA),

    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),

    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),

    plot.title = element_text(size = 18, face = "bold"),
    legend.position = "none"
  )
```

```{r}
diagnosis_fragmentation <- combined %>%
  group_by(PatientDurableKey, PrimaryDiagnosisKey) %>%
  summarise(
    n_primary_specialty = n_distinct(PrimarySpecialty, na.rm = TRUE),
    n_department_specialty = n_distinct(DepartmentSpecialty, na.rm = TRUE),
    .groups = "drop"
  )

diagnosis_fragmentation

journey2 <- read.csv("journey_data.csv")

journey2


df <- diagnosis_fragmentation %>%
  left_join(
    patient_summary %>% select(PatientDurableKey, AnswerText),
    by = "PatientDurableKey"
  ) %>%
  filter(!is.na(AnswerText))

journey2_enriched <- journey2 %>%
  left_join(df,
    by = c("PatientDurableKey", "PrimaryDiagnosisKey")
  )

combined
```

# Primary Specialty
```{r}
patient_level <- df %>%
  group_by(PatientDurableKey, AnswerText) %>%
  summarise(
    avg_primary_specialty = mean(n_primary_specialty, na.rm = TRUE),
    .groups = "drop"
  )

patient_level %>%
  group_by(AnswerText) %>%
  summarise(
    mean_avg_primary_specialty = mean(avg_primary_specialty, na.rm = TRUE),
    sd = sd(avg_primary_specialty, na.rm = TRUE),
    n = n()
  )

ggplot(patient_level, aes(x = AnswerText, y = avg_primary_specialty, fill = AnswerText)) +
  geom_boxplot() +
  labs(
    x = "Transportation Need",
    y = "Avg # of Primary Specialization Types per Diagnosis"
  ) +
    stat_summary(
    fun.data = function(x) data.frame(y = max(x) + 0.3, label = paste0(round(mean(x),3))),
    geom = "text",
    size = 4
  ) +
  theme_classic() +
  scale_fill_manual(values = c("No" = "darkturquoise", "Yes" = "#00BFC4")) +
  theme(legend.position = "none")+
  theme(
    plot.background = element_rect(fill = "#fafafa", color = NA),
    panel.background = element_rect(fill = "#fafafa", color = NA),

    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),

    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),

    plot.title = element_text(size = 18, face = "bold"),
    legend.position = "none"
  )
```

# Department
```{r}
patient_level_dept <- df %>%
  group_by(PatientDurableKey, AnswerText) %>%
  summarise(
    avg_department_specialty = mean(n_department_specialty, na.rm = TRUE),
    .groups = "drop"
  )

patient_level_dept %>%
  group_by(AnswerText) %>%
  summarise(
    mean_avg_department_specialty = mean(avg_department_specialty, na.rm = TRUE),
    sd = sd(avg_department_specialty, na.rm = TRUE),
    n = n()
  )

ggplot(patient_level_dept,
       aes(x = AnswerText, y = avg_department_specialty, fill = AnswerText)) +
  geom_boxplot() +
  
  stat_summary(
    fun.data = function(x) data.frame(y = max(x) + 0.3, label = paste0(round(mean(x),3))),
    geom = "text",
    size = 4
  ) +
  
  labs(
    x = "Transportation Need",
    y = "Avg # of Department Types per Diagnosis",
  ) +
  
  theme_classic() +
  scale_fill_manual(values = c("No" = "darkturquoise", "Yes" = "#00BFC4")) +
  theme(legend.position = "none")+
  theme_classic() +
  scale_fill_manual(values = c("No" = "darkturquoise", "Yes" = "#00BFC4")) +
  theme(legend.position = "none")+
  theme(
    plot.background = element_rect(fill = "#fafafa", color = NA),
    panel.background = element_rect(fill = "#fafafa", color = NA),

    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),

    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),

    plot.title = element_text(size = 18, face = "bold"),
    legend.position = "none"
  )

```



